---
title: "Compulsory Exercise 2: Predicting AirBnB-prices in Rome"
author:
- Helle Villmones Haug
- Hjalmar Jacob Vinje
- Sanna Baug Warholm
date: "`r format(Sys.time(), '%d %B, %Y')`"
header-includes: \usepackage{amsmath}
output:
  # html_document:
  #   toc: no
  #   toc_depth: '2'
  #   df_print: paged
  pdf_document:
    toc: no
    toc_depth: '2'
urlcolor: blue
abstract: "This is the place for your abstract (max 350 words)"
---
  
```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE,tidy=TRUE,message=FALSE,warning=FALSE,strip.white=TRUE,prompt=FALSE,
                      cache=TRUE, size="scriptsize",fig.width=4, fig.height=3,fig.align = "center")
```

```{r,eval=TRUE,echo=FALSE}
library("knitr")
library("rmarkdown")
```

<!--  Etc (load all packages needed). -->





## Introduction: Scope and purpose of your project


The purpose of the project is to use machine learning to predict the price of Airbnb nights in Rome XX TBA. 

The data set used is the rome_weekends.csv file from Kaggle (XX source). It was analyzed by X and X. 

*summarize key findings*

## Descriptive data analysis/statistics



## Methods
### Method 1 (Tree-based?)

### Method 2 (Something more fancy, gradient boosting mabye?)

## Results and interpretation


## Summary