---
title: "Compulsory Exercise 2: Predicting AirBnB-prices in Rome"
author:
- Helle Villmones Haug
- Hjalmar Jacob Vinje
- Sanna Baug Warholm
date: "`r format(Sys.time(), '%d %B, %Y')`"
header-includes: \usepackage{amsmath}
output:
  # html_document:
  #   toc: no
  #   toc_depth: '2'
  #   df_print: paged
  pdf_document:
    toc: no
    toc_depth: '2'
urlcolor: blue
abstract: "The purpose of the project is to use machine learning to predict the price of Airbnb nights in Rome XX TBA. 

The data set used is the rome_weekends.csv file from Kaggle (XX source). It was analyzed by X and X. 

Summarize key findings

Why are findings important/novel?"


---
  
```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE,tidy=TRUE,message=FALSE,warning=FALSE,strip.white=TRUE,prompt=FALSE,
                      cache=TRUE, size="scriptsize",fig.width=4, fig.height=3,fig.align = "center")
```

```{r,eval=TRUE,echo=FALSE}
library("knitr")
library("rmarkdown")
```

<!--  Etc (load all packages needed). -->





## Introduction: Scope and purpose of your project

\textcolor{red}{Briefly introduce the broad idea of the problem or task that you chose and the respective data set that
you use. This could be a classification task (e.g., predicting the species of flowers in the Iris data set) or
a regression task (e.g., predicting the price of a house based on its features). Clearly define the scope of
your project. What specific problem are you trying to solve?
• Describe the source and give a referenc to where the data set is coming from.
• Describe the purpose of your project is some more detail. What are the specific question that you want
to answer in your project? Are you trying to find the best performing method or a good performing
and light method that is easy to use? Who is your audience? Are you trying to discover the relations
between different variables? Are you trying to find important predictors for your classification? Are
you trying to draw some insightful understanding in a particular topic/domain? Importantly: Is the
main purpose inference or prediction?}


In this project, we aim to predict the total rental price of Airbnb listings in Rome during weekends, using a dataset called 'rome_weekends.csv' from Kaggle (https://www.kaggle.com/datasets/thedevastator/airbnb-prices-in-european-cities?select=rome_weekends.csv). This is a regression task, as we are predicting a continuous variable (total rental price) based on various features of the listings.

The dataset is obtained from a public platform, Kaggle, where users can upload, share, and collaborate on data science projects. The data contains information on Airbnb listings in Rome, such as room type, whether the room is shared or private, and if the host is a superhost. Our goal is to utilize this information to predict the total rental price of a given listing and evaluate the performance of two machine learning models.

The purpose of this project is two-fold:

\begin{enumerate}
    \item To compare the performance of two different machine learning techniques: a deep learning model using the Keras library and a Random Forest model. We want to determine which model yields better prediction accuracy and whether there is a trade-off between performance and model complexity.
    \item To uncover relationships between different variables and identify important predictors for the total rental price. This will help us draw insightful understanding in the domain of Airbnb rental prices, which can be beneficial for hosts and guests alike.
\end{enumerate}

Our audience consists of data scientists, machine learning enthusiasts, and Airbnb stakeholders who are interested in gaining a deeper understanding of the factors affecting rental prices and improving price prediction models. The main purpose of this project is to predict the total rental price of Airbnb listings with high accuracy, while also uncovering relationships between variables that can contribute to a better understanding of the underlying factors that influence rental prices. In essence, our project focuses on both prediction and inference, with the goal of achieving high prediction accuracy and gaining insights into the determinants of rental prices.





## Descriptive data analysis/statistics

\textcolor{red}{Conduct descriptive data analysis to get an overview over your data (see this example for inspiration).
Examples:
• Report measures such as mean, median, range, standard deviation, and variance to describe the central
tendency, variability, and distribution of a data set.
• Scatter plots and correlation matrices across different variables and histograms of variables (see this
example).
• Box plots of variables.}

 

## Methods

\textcolor{red}{Describe the methods that you are using in your project and explain in detail how you applied them.
Depending on the task, these could include methods such as linear or logistic regression, decision trees,
random forests, support vector machines etc. You should use several methods for your problem so that
you can compare their performance.
• Explain briefly how each method works, what its strengths and weaknesses are, both in general but
also in the light of your project (how suitable is the method in your case?).
3
• Describe which hyperparameters are optimized for the methods (e.g., a shrinkage factor is a hyperparameter in Lasso).
• Describe clearly how you evaluate the performance of the different models and methods (accuracy, MSE,
misclassification error, CV error,. . . ). Explain how each metric is calculated, and why it is a useful
measure of model performance.
• (optional) Consider and describe potential limitations of the methods and the chosen evaluation metrics.}


### Method 1 A deep learning model using ReLU

ReLU introduces non-linearity in a manner similar to step functions, but it does so using linear segments instead of constant segments.In addition, ReLU uses only one step \textcolor{red}{(can someone confirm?)}. The Keras library functions have features such as 


### Method 2 Random Forest model

## Results and interpretation


## Summary